---
title: "Addiction Treatment as a Sociological Outcome"
author: "Christopher Carbonaro"
date: "December 10th, 2019"
output: 
  beamer_presentation:
    theme: Berkeley
    colortheme: seahorse
    slide_level: 2
    incremental: TRUE
    keep_tex: TRUE
---

# Introduction

## Drug Use in the U.S.

- 60.2% (164.8 Million) of Americans have consumed alcohol, tobacco, or an illicit drug within the past month
- Of these users, 31.9 million used an illicit drug within the past month
- 20% of Americans are estimated to have used an illicit drug within the past year (SAMHSA, 2019, p. 3-8)


## Addiction Treatment in the U.S.

Approximately "21.2 million people aged 12 or older needed substance use treatment" in 2018 (SAMHSA, p. 3).

Unfortunately, while this comprises 7.8% of the U.S. population, only 1.4 received treatment within the past year (p. 3).

## This Project's Focus

Given the disparity between the number of individuals in need of treatment and the number who receive it, we are motivated to ask:

**What sociological factors best predict whether an individual will undergo substance abuse treatment?**


# Literature Review

## Current Research

This project eschews studying addiction as a neurological phenomenon (see Venniro et al., 2017).

Instead, it builds on work which looks at sociological factors which predict addiction (Boyle, Polinsky, & Hser, 2000; Taylor, Caudy, Blasko, & Taxman, 2017; Battjes, Gordon, O’Grady, Kinlock, & Carswell, 2003).

## Current Research (cont.)

Limitations of current research:

- Small sample sizes (Boyle et. al., 2000; Battjes et. al., 2003)
- No use of analytic tools other than multiple regression (*Ibid*.)
- Luongo et. al. (2016) conducted a Canadian study which produced results conflicting with Boyle et. al. (2000) and Battjes et. al. (2003) regarding the importance of:
  - Severity of substance being abused
  - Gender
  - Ethnicity
  - Age
  - ... and Education

## Filling the Research Gap

This project is an attempt to examine the importance of these variables (and others) on a larger scale while 1. looking for interactions between the predictors and 2. using tools other than OLS.

# Methods

## Source of Data

This project uses a survey conducted by the Substance Abuse and Mental Health Services Administration (SAMHSA), namely the 2018 National Survey on Drug Use and Health (NSDUH).

The survey consists of roughly 56,000 participants.

## Predictors in the Data

There are hundreds of variables for each respondant in the NSDUH; this study narrows the focus to roughly 40 predictors.

These include variables like age and gender, variables describing the individual's health insurance coverage, and variables describing their recency/frequency/type of drug use.

## Logistic Regression

Our first tool for examining the data is regularized regression. We use this as a baseline against which to compare our second tool.

Here, we use the elastic net to assess the importance of each variable and avoid producing an overcomplicated model.

## Logistic Regression (cont.)

$$log\frac{Pr(G = 2|X=x)}{Pr(G=1|X=x)}=\beta_0 + \beta^Tx$$

## Logistic Regression (cont.)

$${\widehat{\beta}}^{\text{elastic net}} = \text{argmin}\frac{1}{2}\sum_{i = 1}^{N}{(y_{i} - \beta_{0} - \sum_{j = 1}^{p}x_{\text{ij}}\beta_{j})^{2}}$$
$$ + \lambda\sum_{j = 1}^{p}{{(\alpha\beta_{j}^{2} + (1 - \alpha)|\beta}_{j}|)}$$
  
## Boosting

One of the biggest problems with linear regression is we have a comparatively small number of positive cases (only around 5%).

This makes classifying positive cases difficult.

## Boosting (cont.)

Boosting is a helpful solution here, since it will raise the relative importance of misclassified observations for each subsequent tree.

We can also use partial dependence plots to look for interactions between important terms.

## Boosting (cont.)

From Witten, Hastie, & Tibshirani (2017, p. 323):

1. Set ${\widehat{f}(x) = 0}$ and $r_i = y_i$ for all $i$ in the training set.
2. For b = 1, 2, ... , B, repeat:
  a. Fit a tree $\widehat{f}^b$ with d splits to the training data
  b. Update $\widehat{f}$ by adding in a shrunken version of the new tree: $$\widehat{f}(x) \leftarrow \widehat{f}(x) + \lambda \widehat{f}^b(x)$$
  c. Update the residuals: $$r_i \leftarrow r_i - \lambda \widehat{f}^b(x_i)$$
3. Output the boosted model: $$\widehat{f}(x) = \sum_{b=1}^{B}{\lambda \widehat{f}^b(x)}$$

# Results

## OLS Results

```{r loadCleanData, echo=FALSE, message=F, warning=F}
require(pacman)
p_load(tidyverse,
       caret,
       psych,
       mice,
       glmnet,
       gbm,
       pdp,
       partykit,
       party,
       mclust,
       mixtools,
       e1071)

load("C:/Users/clcar/Dropbox/College/Senior Year/Machine_Learning_for_Social_Science/Final_Project/NSDUH-2018/NSDUH_2018.RData")

nsduh <- PUF2018_100819

data_filtered <- nsduh %>%
  select(TXEVRRCVD2, # Received Treatment OUTCOME
         irsex, # Gender
         irmarit, # Marital Status
         IREDUHIGHST2, # Level of Education
         AGE2, # Age
         service, # Military Service
         HEALTH2, # Self-Assessment of Health
         sexatract, # Sexual Orientation
         NEWRACE2, # Ethnicity
         irwrkstat, # Employment Status
         IRHHSIZ2, # Number of People in Household
         IRKI17_2, # Number of children under 18
         irmcdchp, # Has Medicaid/CHIP
         irmedicr, # Has Medicare
         irprvhlt, # Has Private Health Insurance,
         IRINSUR4, # Has any kind of Health Insurance,
         POVERTY3, # Living in Poverty
         ircigrc, # Cigarette Recency
         iralcrc, # Alcohol Recency
         irmjrc, # Marijuana Recency
         ircocrc, # Cocaine Recency
         ircrkrc, # Crack Recency
         irherrc, # Heroin Recency
         irhallucrec, # Hallucinogen Recency
         irecstmorec, # Ecstacy Recency
         irketminrec, # Ketamine Recency
         irdamtfxrec, # DMT/AMT/FOXY Recency
         irsalviarec, # Salvia Recency
         irinhalrec, # Inhalant Recency
         irmethamrec, # Methamphetamine Recency
         irpnrnmrec, # Pain Reliever Misuse Recency
         irtrqnmrec, # Tranquilizer Misuse Recency
         irstmnmrec, # Stimulate Misuse Recency
         irsednmrec, # Sedative Misuse Recency
         iralcfy, # Alcohol Frequency pst. yr.
         irmjfy, # Marijuana Frequency pst. yr
         ircocfy, # Cocaine Frequency pst. yr.
         ircrkfy, # Crack Frequency pst. yr.
         irherfy, # Heroin Frequency pst. yr.
         irhallucyfq, # Hallucinogen Frequency pst. yr.
         irinhalyfq, # Inhalant Frequency pst. yr.
         irmethamyfq # Meth Frequency pst. yr.
         ) %>%
  mutate(TXEVRRCVD2 = factor(TXEVRRCVD2))

elastic_trainControl <- trainControl(method = "cv",
                                     number = 10)

elastic <- train(TXEVRRCVD2 ~ .,
                 data = data_filtered,
                 method = "glmnet",
                 trControl = elastic_trainControl,
                 tuneGrid = expand.grid(
                   .alpha = 0.5,
                   .lambda = seq(0, 1, by = 0.05)
                 ),
                 na.action = na.omit)

### y_elastic <- data_filtered$TXEVRRCVD2
### x_elastic <- data_filtered %>%
###   select(-TXEVRRCVD2) %>%
###   as.matrix.data.frame()
### 
### elastic2 <- cv.glmnet(x=x_elastic,
###                       y=y_elastic,
###                       family="binomial")

```

```{r}
knitr::kable(elastic$results)
```

***

```{r, comment=""}
summary(elastic$results)
```

***

```{r}
plot(elastic)
```

## Boosting Results

```{r}

```


# Next Steps



# Works Cited

## Resources

Battjes, R. J., Gordon, M. S., Ogrady, K. E., Kinlock, T. W., & Carswell, M. A. (2003). Factors that Predict Adolescent Motivation for Substance Abuse Treatment. *Journal of Substance Abuse Treatment*, 24(3), 221–232. doi: 10.1016/s0740-5472(03)00022-9

Boyle, K., Polinsky, M. L., & Hser, Y.-I. (2000). Resistance to Drug Abuse Treatment: A Comparison of Drug Users Who Accept or Decline Treatment Referral Assessment. *Journal of Drug Issues*, 30(3), 555–574. doi: 10.1177/002204260003000304

Cohen, J., Ilumoka, A., & Salehi, I. (2015). Neural Network-Based Drug Abuse Treatment Optimization. *Procedia Computer Science*, 61, 454–459. doi: 10.1016/j.procs.2015.09.186

## Resources (cont.)

Hastie, T., Tibshirani, R., & Friedman, J. (2009). *The Elements of Statistical Learning Data Mining, Inference, and Prediction, Second Edition* (2nd ed.). New York, NY: Springer New York.

James, G., Witten, D., Hastie, T., & Tibshirani, R. (2017). *An Introduction to Statistical Learning: With Applications in R*. New York: Springer.

## Resources (cont.)

The Substance Abuse and Mental Health Services Administration. (2019). The Key Substance Use and Mental Health Indicators in the United States: Results from the 2018 National Survey on Drug Use and Health. Retrieved from https://www.samhsa.gov/data/sites/default/files/cbhsq-reports/NSDUHNationalFindingsReport2018/NSDUHNationalFindingsReport2018.pdf

Luongo, N. M., Dong, H., Kerr, T. H., Milloy, M. J. S., Hayashi, K., & Richardson, L. A. (2017). Income Generation and Attitudes Towards Addiction Treatment Among People who use Illicit Drugs in a Canadian Setting. *Addictive Behaviors*, 64, 159–164. doi: 10.1016/j.addbeh.2016.08.041

## Resources (cont.)

Venniro, M., Caprioli, D., Zhang, M., Whitaker, L. R., Zhang, S., Warren, B. L., … Shaham, Y. (2017). The Anterior Insular Cortex→Central Amygdala Glutamatergic Pathway Is Critical to Relapse after Contingency Management. *Neuron*, 96(2). doi: 10.1016/j.neuron.2017.09.024

Yarkoni, T., & Westfall, J. (2017). Choosing Prediction Over Explanation in Psychology: Lessons From Machine Learning. *Perspectives on Psychological Science*, 12(6), 1100–1122. doi: 10.1177/1745691617693393

